---
title: "Data preparation"
author: "Marieke Vinkenoog"
date: "18/06/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(lubridate)
library(dummies)
```

# Load data
Data has been preprocessed and merged. Only the Time variable is missing, which
we will include from the raw data file.

```{r warning=FALSE}
data <- readRDS("X:/202003 Ferritine en omgeving/donordata_with_temps_env.RDS")
ddata_raw <- read.delim("X:/202003 Ferritine en omgeving/Raw data/donations_201710_201912.dat")
ddata_raw$Donatiedatum <- as.Date(ddata_raw$Donatiedatum, format="%m/%d/%Y")
data <- merge(data, ddata_raw[c("KeyID", "Donatiedatum", "Donatie_Tijd_Start", "Hb")],
              by.x=c("DonorID", "Date"), by.y=c("KeyID", "Donatiedatum"))
colnames(data)[29] <- "Time"
rm(ddata_raw)
```

# Variables
The data set contains 30 variables and 140713 observations. Each observation is 
a ferritin measurement from a successful blood donation (DonType V) or from
a new donor intake (DonType N). Each donor only occurs once in the data set.

Numerical variables: Age, DaysSincePrev, NumBefore, BloodVolume, BMI, 
                     AvgTemp (1/7/30), MinTemp (1/7/30), MaxTemp (1/7/30),
                     SES, PopDensity, O3, PM25, M10, EC, NO2
Binary variables   : Sex, Duffy
Datetime variables : Date, Time

Numerical variables are ready to go.
Binary variables must be recoded as 0/1 (currently factors).
Datetime variables need to be recoded: time will be represented as a float
and date will be represented by two variables (sin&cos) to account for
the cyclical nature of the variable. (Models don't know that December 31st and
January 1st are close otherwise.) Year is disregarded. Recoded date&time can
be used as numerical variables in the model.

```{r warning=FALSE}
data$Sex <- dummy(data$Sex)[, 1]       # 0 for male, 1 for female
data$Duffy <- dummy(data$Duffy)[, 1]   # 0 for yes, 1 for no
data$Time <- as.numeric(substr(data$Time, 0, 2)) + as.numeric(substr(data$Time, 4, 5)) / 60
data[, c("Date_sin", "Date_cos")] <- cyclic_encoding(data$Date, "year")
```
# Inspect variables for outliers

Temperature variables have been checked for outliers during preprocessing,
so we can leave those be.

```{r}
summary(data[c("Age", "Ferritin", "DaysSincePrev", "NumBefore", "BloodVolume",
               "BMI", "SES", "PopDensity", "O3", "PM25", "PM10", "EC", "NO2")])
```
Suspicious variables: Ferritin (max), DaysSincePrev (min), BloodVolume (min), BMI (min&max).
We remove the following observations:
- Ferritin > 350 (normal range for men goes up to 336 ng/mL, higher values can indicate underlying conditions)
- DaysSincePrev < 56 (minimum inter-donation interval for men)
- BloodVolume < 3 (minimum 'normal' BV for a 50 kg woman (minimum weight for blood donors) is 3.2L)
- BMI < 16 or > 50 (lower or higher values are likely due to incorrect height/weight recorded)

Also remove the 25 observations that have NA values for all environmental variables.

```{r}
data <- data[data$Ferritin <= 350, ]              # 913 observations removed
data <- data[(is.na(data$DaysSincePrev)) | 
             (data$DaysSincePrev >= 56), ]        # 1528 observations removed 
data <- data[!(is.na(data$BloodVolume)) & 
              (data$BloodVolume > 3.2), ]         # 189 observations removed
data <- data[!(is.na(data$BMI)) & 
              (data$BMI > 16) & 
              (data$BMI < 50), ]                  # 46 observations removed
data <- data[!(is.na(data$NO2)), ]                # 25 observations removed
```

```{r}
summary(data[c("Age", "Ferritin", "DaysSincePrev", "NumBefore", "BloodVolume",
               "BMI", "SES", "PopDensity", "O3", "PM25", "PM10", "EC", "NO2")])
```
```{r}
saveRDS(data, "X:/202003 Ferritine en omgeving/donordata_with_temps_env_cleaned.RDS")
```

